\acknowledgementtext{%
	I would like to express my sincere gratitude to my supervisor, Prof. Stevan Dubljevic, for his continuous guidance and support throughout my Ph.D. studies. His insights and encouragement have shaped the direction of this research, and I am grateful for the opportunity to pursue my doctoral work under his supervision. I would also like to thank the members of my examination committee for their time, feedback, and thoughtful evaluation of my thesis.

	I gratefully acknowledge the financial support provided by the Natural Sciences and Engineering Research Council (NSERC) of Canada, as well as the Department of Chemical and Materials Engineering at the University of Alberta. I would also like to thank Daniel Aldrich for his development and maintenance of the \texttt{ualberta-thesis} \LaTeX{} template, which greatly facilitated the preparation of this document.

	I am especially thankful to my colleagues and collaborators in the Distributed Parameter Systems Lab for their support, discussions, and shared experiences. In particular, I would like to acknowledge Guilherme Ozorio Cassol, Lu Zhang, Junyao Xie, Mojtaba Izadi, Jukka-Pekka Humaloja, and Farzin Sadelari for their valuable contributions during different phases of this work. Among all other peers who have been part of my journey, I would like to extend my gratitude to Dr. Yousef Salehi and Amir Mohseni, whose companionship and support have been invaluable inside and outside the office.

	I am thankful to my best friend Parsa, who stood by me over the years—no matter how far away he was—offering encouragement and companionship through the inevitable highs and lows of academic life. To my lovely sister, I am forever grateful for your kindness and for your presence, which has been a quiet source of strength. To my parents, thank you for the foundation you built and the sacrifices you made. And finally, to my wife—thank you for being my constant companion, my greatest support, and the person who reminded me what truly matters beyond these pages.
}