\section{Conclusion} \label{sec:conclusion}

This work establishes a delay-aware, late-lumped framework for output-based estimation and constrained control of a non-isothermal axial dispersion tubular reactor with recycle. By modeling the state delay as a transport PDE, the recycle stream is incorporated into the system without approximation, preserving the infinite-dimensional structure of the coupled mass--energy dynamics. The proposed Cayley--Tustin discretization enables structure-preserving transition to a discrete-time setting, while the use of closed-form resolvent operators eliminates the need for spatial discretization or model reduction.

Full-state MPC is shown to effectively stabilize an unstable setup while enforcing input constraints, confirming its viability for delay-affected infinite-dimensional systems. The MHE--MPC framework enables constrained output feedback control using only partial measurements; i.e. essential for infinite-dimensional systems where full-state information is generally inaccessible. The optimization-based structure of the integrated MHE--MPC framework naturally accommodates constraints while the fact that it operates in discrete-time setting makes it well suited for digital implementation, especially in the presence of measurement noise or model inaccuracies. However, unlike observer-based designs that allow explicit pole placement, MHE does not permit direct shaping of the estimation error dynamics. As a result, there is no guarantee that the estimator converges faster than the controller acts---a requirement for closed-loop stability in output-feedback settings. To avoid this limitation, the integrated MHE--MPC strategy is applied to a nominally stable system, where it reconstructs distributed states under process and measurement noise and stabilizes the system within an optimal framework, while successfully accommodating constraints throughout closed-loop operation.

To conclude, the obtained modeling framework along with the proposed MHE--MPC architecture provide a modular foundation for infinite-dimensional estimation and control of diffusion--convection--reaction systems with state delays, as encountered in advanced chemical engineering applications.